Automated Assessment of Residual Plots with Computer Vision Models

### 2025-06-13 paper.Rmd
Combine model training with data generation into one section

In this paper, we develop computer vision models and integrate them into the residual plots diagnostics workflow, addressing the need for an automated visual inference. The paper is structured as follows. Section \ref{sec-model-specifications} discusses various specifications of the computer vision models. Section \ref{sec-model-distance-between-residual-plots} defines the distance measure used to detect model violations, while Section \ref{sec-model-distance-estimation} explains how the computer vision models estimate this distance measure. Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance, and Section \ref{sec-model-violations-index} introduces a Model Violations Index, which offers a quicker and more convenient assessment. Sections \ref{sec-model-data-generation}, \ref{sec-model-architecture}, and  \ref{sec-model-training} detail the data preparation, model architecture, and training process, respectively. The results are presented in Section \ref{sec-model-results}. Example dataset applications are discussed in Section \ref{sec-examples}. Finally, we conclude with a discussion of our findings and propose ideas for future research directions.

@@

In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot. This is further explained in @sec-model-distance-between-residual-plots. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

@@

# Data Generation {#sec-model-data-generation}

The simulated data incorporated three common types of residual departures: non-linearity, heteroskedasticity, and non-normality. These were introduced by fitting a standard simple linear regression model to data generated from a more flexible model based on Hermite polynomials, multiple predictors, and varying distributions in both predictors and error terms. A comprehensive grid of parameter combinations was explored to generate a wide range of residual patterns. Importantly, the simulation included scenarios with multiple violations occurring simultaneously.

To ensure uniform coverage across the difficulty scale of the target variable $D$, a bucket sampling scheme was used to create a balanced dataset. Samples were iteratively simulated and accepted into one of 50 buckets, each representing a distinct range of $D$.

The mathematical formulations of the synthetic data model, parameter specifications, supporting tables and example figures of visual patterns of simulated residuals are provided in the Appendix.

@@

# Model Training {#sec-model-training}

The computer vision models were trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras], with hyperparameters optimized via Bayesian tuning in `KerasTuner` [@omalley2019kerastuner] to minimize validation RMSE. The tuning process involved 100 trials, early stopping, and consideration of factors such as dropout, batch normalization, input resolution, and auxiliary inputs. Details of the hyperparameter domains, optimization strategy, and final selected values are provided in Appendix.

@@

In @li2024plot, the residual plots are simulated from a data generating process which is a special case of Equation \ref{eq:data-sim}. The main characteristic is the model violations are introduced separately, meaning non-linearity and heteroskedasticity will not co-exist in one lineup but assigned uniformly to all lineups. Additionally, non-normality and multiple predictors are not considered in the experimental design.

>>>>

In this paper, we develop computer vision models and integrate them into the residual plots diagnostics workflow, addressing the need for an automated visual inference. The paper is structured as follows. Section \ref{sec-model-specifications} discusses various specifications of the computer vision models. Section \ref{sec-model-distance-between-residual-plots} defines the distance measure used to detect model violations, while Section \ref{sec-model-distance-estimation} explains how the computer vision models estimate this distance measure. Section \ref{sec-model-statistical-testing} covers the statistical tests based on the estimated distance, and Section \ref{sec-model-violations-index} introduces a Model Violations Index, which offers a quicker and more convenient assessment. Sections \ref{sec-model-architecture} and \ref{sec-model-data-generation} describe the model architecture and the corresponding data generation and training process, respectively. The results are presented in Section \ref{sec-model-results}. Example dataset applications are discussed in Section \ref{sec-examples}. Finally, we conclude with a discussion of our findings and propose ideas for future research directions.

@@

In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot. This is further explained in \ref{sec-model-distance-estimation}. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

@@

@@

# Data Generation and Model Training {#sec-model-data-generation}

The simulated data incorporated three common types of residual departures: non-linearity, heteroskedasticity, and non-normality. These were introduced by fitting a standard simple linear regression model to data generated from a more flexible model based on Hermite polynomials [@hermite1864nouveau], multiple predictors, and varying distributions in both predictors and error terms. A comprehensive grid of parameter combinations was explored to generate a wide range of residual patterns. Importantly, the simulation included scenarios with multiple violations occurring simultaneously. To ensure uniform coverage across the difficulty scale of the target variable $D$, a bucket sampling scheme was used to create a balanced dataset. Samples were iteratively simulated and accepted into one of 50 buckets, each representing a distinct range of $D$.

Model training was conducted on the MASSIVE M3 high-performance computing platform [@goscinski2014multi] using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras]. Hyperparameters were optimized via Bayesian tuning with KerasTuner [@omalley2019kerastuner], minimizing validation RMSE across 100 trials. The tuning process included early stopping and considered dropout, batch normalization, input resolution, and auxiliary inputs.

Further details, including mathematical formulations of the simulation model, parameter specifications, sampling scheme, example residual plots and hyperparameter tuning configuration, are provided in the Appendix.

@@

In @li2024plot, the residual plots are simulated from a data generating process that corresponds to a special case of the synthetic data model used in this study. A key feature of the design is that model violations are introduced independently, meaning non-linearity and heteroskedasticity do not co-exist within a single lineup but are instead assigned uniformly across different lineups. Moreover, the experimental design does not account for non-normality or multiple predictors.

### 2025-06-13 paper.tex
Combine model training with data generation into one section

In this study, we chose to define and use a distance between a true
residual plot and a theoretically ``good'' residual plot. This is
further explained in \citet{sec-model-distance-between-residual-plots}.
\citet{vo2016localizing} have also demonstrated that defining a proper
distance between images can enhance the matching accuracy in image
search compared to a binary outcome model.

@@

\section{Data Generation}\label{sec-model-data-generation}

The simulated data incorporated three common types of residual
departures: non-linearity, heteroskedasticity, and non-normality. These
were introduced by fitting a standard simple linear regression model to
data generated from a more flexible model based on Hermite polynomials,
multiple predictors, and varying distributions in both predictors and
error terms. A comprehensive grid of parameter combinations was explored
to generate a wide range of residual patterns. Importantly, the
simulation included scenarios with multiple violations occurring
simultaneously.

To ensure uniform coverage across the difficulty scale of the target
variable \(D\), a bucket sampling scheme was used to create a balanced
dataset. Samples were iteratively simulated and accepted into one of 50
buckets, each representing a distinct range of \(D\).

The mathematical formulations of the synthetic data model, parameter
specifications, supporting tables and example figures of visual patterns
of simulated residuals are provided in the Appendix.

@@

\section{Model Training}\label{sec-model-training}

The computer vision models were trained on the MASSIVE M3
high-performance computing platform \citep{goscinski2014multi}, using
TensorFlow \citep{abadi2016tensorflow} and Keras
\citep{chollet2015keras}, with hyperparameters optimized via Bayesian
tuning in \texttt{KerasTuner} \citep{omalley2019kerastuner} to minimize
validation RMSE. The tuning process involved 100 trials, early stopping,
and consideration of factors such as dropout, batch normalization, input
resolution, and auxiliary inputs. Details of the hyperparameter domains,
optimization strategy, and final selected values are provided in
Appendix.

@@

In \citet{li2024plot}, the residual plots are simulated from a data
generating process which is a special case of Equation
\ref{eq:data-sim}. The main characteristic is the model violations are
introduced separately, meaning non-linearity and heteroskedasticity will
not co-exist in one lineup but assigned uniformly to all lineups.
Additionally, non-normality and multiple predictors are not considered
in the experimental design.

>>>>

In this study, we chose to define and use a distance between a true
residual plot and a theoretically ``good'' residual plot. This is
further explained in \ref{sec-model-distance-estimation}.
\citet{vo2016localizing} have also demonstrated that defining a proper
distance between images can enhance the matching accuracy in image
search compared to a binary outcome model.

@@

@@

\section{Data Generation and Model
Training}\label{sec-model-data-generation}

The simulated data incorporated three common types of residual
departures: non-linearity, heteroskedasticity, and non-normality. These
were introduced by fitting a standard simple linear regression model to
data generated from a more flexible model based on Hermite polynomials
\citep{hermite1864nouveau}, multiple predictors, and varying
distributions in both predictors and error terms. A comprehensive grid
of parameter combinations was explored to generate a wide range of
residual patterns. Importantly, the simulation included scenarios with
multiple violations occurring simultaneously. To ensure uniform coverage
across the difficulty scale of the target variable \(D\), a bucket
sampling scheme was used to create a balanced dataset. Samples were
iteratively simulated and accepted into one of 50 buckets, each
representing a distinct range of \(D\).

Model training was conducted on the MASSIVE M3 high-performance
computing platform \citep{goscinski2014multi} using TensorFlow
\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}.
Hyperparameters were optimized via Bayesian tuning with KerasTuner
\citep{omalley2019kerastuner}, minimizing validation RMSE across 100
trials. The tuning process included early stopping and considered
dropout, batch normalization, input resolution, and auxiliary inputs.

Further details, including mathematical formulations of the simulation
model, parameter specifications, sampling scheme, example residual plots
and hyperparameter tuning configuration, are provided in the Appendix.

@@

In \citet{li2024plot}, the residual plots are simulated from a data
generating process that corresponds to a special case of the synthetic
data model used in this study. A key feature of the design is that model
violations are introduced independently, meaning non-linearity and
heteroskedasticity do not co-exist within a single lineup but are
instead assigned uniformly across different lineups. Moreover, the
experimental design does not account for non-normality or multiple
predictors.
